---
title: "Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(ggplot2)
library(tidyverse)
library(gridExtra)
```


This vignette describes the two latent variable models implemented in
`lblvm`: the **bounded-continuous response model** and the **discretized
response model (DRM)**. Both models use the same logit-beta framework, but
differ in how the response is observed.

```{r, fig.cap="**Figure 1** Bounded-continuous response model in Panel (a) and discretized response model in Panel (b). The black solid lines in the two panels are identical and are the $\\mu$-curves. The grey curves represent the probability densities of continuous item responses in Panel (a) and those of internal judgment continuum in Panel (b). The grey line segments, located exactly on the $\\mu$-curve, indicate the means of the densities. The relative width of the color bands in Panel (b) can be estimated in the discretized response model.", fig.width=12, fig.height=5, out.width="100%", fig.align='left', echo=FALSE, warning=FALSE, message=FALSE}
P <- function(theta,a=1,b,c=0){
  c+(1-c)*(1/(1+exp(-a*(theta-b))))
}
nu <- 10
apar <- 1
bpar <- 0
item_par <- c(apar,bpar,nu)#x$par_est[item.number,]

# Continuous IRT & DRM

probs <- P(theta = seq(-6,6,0.1),
           a = item_par[1],
           b = item_par[2])
lower.ci <- stats::qbeta(.025, item_par[3]*probs, (1-probs)*item_par[3])
upper.ci <- stats::qbeta(.975, item_par[3]*probs, (1-probs)*item_par[3])
p_df <- data.frame(response = c(probs, lower.ci, upper.ci),
                   theta = rep(seq(-6,6,0.1), 3),
                   type = rep(c("Mean", "C.I. (2.5%)", "C.I. (97.5%)"), each = 121))

ppp <-
  ggplot2::ggplot()+
  ggplot2::geom_line(mapping = aes(x=p_df[p_df$type=="Mean","theta"],y=p_df[p_df$type=="Mean","response"]), linewidth=1.5)+
  ggplot2::lims(y=c(0,1))

for(i in c(-2, -1, 0, 1, 2)){
  prob_irt <- P(i, apar,bpar)
  alpha <- prob_irt*nu
  beta <- (1-prob_irt)*nu
  xxx <- seq(0.00000001, 0.99999999, 0.0001)
  ddf <- data.frame(x = xxx, y = i-(dbeta(xxx, alpha, beta))*sqrt(prob_irt*(1-prob_irt))/2)
  ddf2 <- data.frame(x=c(i-(dbeta(prob_irt, alpha, beta))*sqrt(prob_irt*(1-prob_irt))/2, i), y=rep(prob_irt,2))

  ppp <- ppp+
    geom_path(data = ddf %>% arrange(x), aes(x = y, y = x),
              linewidth=1, alpha = 1, inherit.aes = FALSE, color="darkgrey", linetype = "solid")+
    geom_path(data = ddf2, aes(x = x, y = y),
              linewidth=1, alpha = 1, inherit.aes = FALSE, color="darkgrey", linetype = "solid")
}

ppp_cont <- ppp+
  geom_hline(yintercept = 0, color="grey", size=1)+
  geom_hline(yintercept = 1, color="grey", size=1)+
  annotate("segment",
           x = c(-2, -1, 0, 1, 2),
           xend = c(-2, -1, 0, 1, 2),
           y=0,
           yend=1,
           color="grey",
           size=1,
           linetype="dashed")+
  scale_x_continuous(breaks= -3:3)+
  coord_cartesian(xlim= c(-3,3))+
  scale_y_continuous(breaks= (0:5)/5,limits = c(0,1))+
  ggplot2::labs(
    title = "(a) Bounded Continuous Response",#sprintf("Item %i", item.number),
    x = expression(theta),
    y = "response")+
  ggplot2::theme_bw()

ppp_drm <- ppp+
  geom_hline(yintercept = 0, color="grey", size=1)+
  geom_hline(yintercept = 1, color="grey", size=1)+
  annotate("segment",
           x = c(-2, -1, 0, 1, 2),
           xend = c(-2, -1, 0, 1, 2),
           y=0,
           yend=1,
           color="grey",
           size=1,
           linetype="dashed")+
  geom_rect(aes(xmin = -Inf, xmax = Inf, ymin = 0, ymax = 0.2),
            fill = "red", alpha = 0.2, inherit.aes = FALSE)+
  geom_rect(aes(xmin = -Inf, xmax = Inf, ymin = 0.2, ymax = 0.4),
            fill = "yellow", alpha = 0.2, inherit.aes = FALSE)+
  geom_rect(aes(xmin = -Inf, xmax = Inf, ymin = 0.4, ymax = 0.6),
            fill = "green", alpha = 0.2, inherit.aes = FALSE)+
  geom_rect(aes(xmin = -Inf, xmax = Inf, ymin = 0.6, ymax = 0.8),
            fill = "dodgerblue3", alpha = 0.2, inherit.aes = FALSE)+
  geom_rect(aes(xmin = -Inf, xmax = Inf, ymin = 0.8, ymax = 1),
            fill = "purple4", alpha = 0.2, inherit.aes = FALSE)+
  annotate("text",
           x = 3,
           y = c(0.1, 0.3, 0.5, 0.7, 0.9),
           label = c("Strongly disagree",
                     "Disagree",
                     "Neutral",
                     "Agree",
                     "Strongly agree"),
           hjust = 1,
           size = 5) +
  scale_x_continuous(breaks= -3:3)+
  coord_cartesian(xlim= c(-3,3))+
  scale_y_continuous(breaks= (0:5)/5,limits = c(0,1))+
  ggplot2::labs(
    title = "(b) Ordered Categorical Response",#sprintf("Item %i", item.number),
    x = expression(theta),
    y = "internal judgment scale")+
  ggplot2::theme_bw()

grid.arrange(grobs=list(ppp_cont,ppp_drm), ncol=2)
```

## Bounded-continuous response model

The bounded-continuous response model is designed for responses that take
values between 0 and 1. Let $x_{ij}$ denote the response of person $i$ to
item $j$, and let $\theta_i$ denote the person's latent trait. The expected
response is modeled as

$$
\operatorname{logit}(\mu_{ij})
=
a_j\theta_i+c_j,
$$

where

$$
\mu_{ij}
=
E(x_{ij}\mid\theta_i)
=
\frac{\exp(a_j\theta_i+c_j)}
{1+\exp(a_j\theta_i+c_j)}.
$$

Here, $a_j$ is the item slope and $c_j$ is the item intercept. The slope
determines how rapidly the expected response changes as the latent trait
changes, whereas the intercept determines the location of the expected
response curve. The latent trait can be multidimensional, and in such cases,
$a$ and $\theta$ become $D$-dimensional vectors and $a_j\theta_i$ in the 
above equation is substituted with an inner product 
$\mathbf{a}_j\cdot\mathbf{\theta}_i=\sum_{d=1}^{D}a_{jd}\theta_{id}$.

The response is then assumed to follow a beta distribution,

$$
x_{ij}\mid\theta_i
\sim
\operatorname{Beta}
\left(
\mu_{ij}\nu_j,
(1-\mu_{ij})\nu_j
\right),
$$

where $\nu_j>0$ is an item-specific precision parameter. Under this
parameterization,

$$
\operatorname{Var}(x_{ij}\mid\theta_i)
=
\frac{\mu_{ij}(1-\mu_{ij})}{\nu_j+1}.
$$

The precision parameter controls how concentrated the responses are around
their expected value. Larger values of $\nu_j$ correspond to greater
precision and, therefore, less residual variation.

The bounded-continuous model can therefore be understood in two parts. The
logit equation describes the relationship between the latent trait and the
expected response, while the beta distribution describes how individual
responses vary around that expectation.

## Discretized response model

The **discretized response model (DRM)** extends the bounded-continuous
model to ordered categorical responses. The key assumption is that an
ordered categorical response arises by discretizing an underlying
bounded-continuous judgment.

For example, consider an item with five response categories:

```text
Strongly disagree | Disagree | Neutral | Agree | Strongly agree
```

DRM assumes that these categories correspond to five consecutive intervals
on an underlying continuum from 0 to 1. Let the boundaries of these
intervals be

$$
0=T_0<T_1<T_2<T_3<T_4<T_5=1.
$$

The underlying continuous response follows the same beta distribution as
the bounded-continuous model. Therefore, for category $z$, the probability
of observing that category is the probability that the underlying response
falls within its corresponding interval:

$$
P(U_{ij}=z\mid\theta_i)
=
\int_{T_{z-1}}^{T_z}
\operatorname{Beta}
\left(x \mid
\mu_{ij}\nu_j,
(1-\mu_{ij})\nu_j
\right)\,dx,
$$

where $\operatorname{Beta}\left(x \mid \mu_{ij}\nu_j,(1-\mu_{ij})\nu_j\right)$ is the beta density implied by the
bounded-continuous model.

For example, the probability of the fourth category is

$$
P(U_{ij}=4\mid\theta_i)
=
\int_{T_3}^{T_4}
\operatorname{Beta}
\left(x \mid
\mu_{ij}\nu_j,
(1-\mu_{ij})\nu_j
\right)\,dx.
$$

Thus, the probability of a category is determined by both the underlying
judgment distribution and the interval assigned to that category.


## Relationship between the two models

The two models share the same underlying logit-beta response process. Their
difference is the observation mechanism.

For the bounded-continuous response model, the underlying response is
observed directly:

$$
\theta_i
\longrightarrow
\mu_{ij}
\longrightarrow
x_{ij}.
$$

For DRM, the same underlying response is discretized into ordered
categories:

$$
\theta_i
\longrightarrow
\mu_{ij}
\longrightarrow
x_{ij}
\longrightarrow
U_{ij}.
$$

The first two stages are identical. The difference is that DRM introduces
thresholds that divide the bounded continuum into intervals, and the
observed category indicates which interval contains the underlying
judgment.

Consequently, DRM allows the latent-trait effects and the response
categorization process to be represented separately: the latent trait
determines the location and precision of the underlying judgment
distribution, while the thresholds determine how that judgment is
translated into an ordered response.

